---
title: Fitting and interpreting logistic regression models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
library(knitr)
library(tidyverse)

dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

### Last time: Dengue data

.large[
**Data:** Data on 5720 Vietnamese children, admitted to the hospital with possible dengue fever. Variables include:

* *Sex*: patient's sex (female or male)
* *Age*: patient's age (in years)
* *WBC*: white blood cell count
* *PLT*: platelet count
* other diagnostic variables...
* *Dengue*: whether the patient has dengue (0 = no, 1 = yes)
]

---

### Logistic regression model

.large[
$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 WBC_i$$
]

---

### Logistic regression model

.large[
$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 WBC_i$$

.question[
Why is there no noise term $\varepsilon_i$ in the logistic regression model? Discuss for 1--2 minutes with your neighbor, then we will discuss as a class.
]
]

---

### Fitting the logistic regression model

.large[

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 WBC_i$$

```{r, eval=F}
m1 <- glm(Dengue ~ WBC, data = dengue, 
          family = binomial)
summary(m1)
```
]

---

### Fitting the logistic regression model

.large[

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 WBC_i$$

```{r, output.lines=9:13}
m1 <- glm(Dengue ~ WBC, data = dengue, 
          family = binomial)
summary(m1)
```
]

---

### Making predictions

.large[
$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{\widehat{p}_i}{1 - \widehat{p}_i}\right) = 1.737 - 0.361 \ WBC_i$$

.question[
Work in groups of 2-3 for 5 minutes on the following questions:

* What is the predicted odds of dengue for a patient with a WBC of 10?
* For a patient with a WBC of 10, is the predicted probability of dengue $> 0.5$, $< 0.5$, or $= 0.5$?
* What is the predicted *probability* of dengue for a patient with a WBC of 10?
]
]

---

### Shape of the regression curve

.large[
.pull-left[
$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 \ X_i \hspace{0.5cm}$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = -1 + seq(-5, 6, length.out=100)) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 2.5) +
  theme_bw() +
  labs(x = "x", y = "Log odds") +
  theme(text = element_text(size = 30))
```
]

.pull-right[
$p_i = \dfrac{e^{\beta_0 + \beta_1 \ X_i}}{1 + e^{\beta_0 + \beta_1 \ X_i}}$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

expit <- function(p){
  return(exp(p)/(1 + exp(p)))
}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 2.5) +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  theme(text = element_text(size = 30))
```
]
]

---

### Plotting the fitted model for dengue data

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
dengue %>%
  ggplot(aes(x = WBC, y = Dengue)) +
  geom_point(size=2) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              lwd = 1.2, se=F) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(y = "P(Dengue = 1 | WBC)")
```

---

### Shape of the regression curve

.large[
How does the shape of the fitted logistic regression depend on $\beta_0$ and $\beta_1$?

.pull-left[
$p_i = \dfrac{\exp\{\beta_0 +  X_i \}}{1 + \exp\{\beta_0 + X_i \}} \hspace{0.5cm}$ for $\beta_0 = -3, -1, 1$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

expit <- function(p){
  return(exp(p)/(1 + exp(p)))
}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + seq(-5, 6, length.out=100)),
           y2 = expit(-3 + seq(-5, 6, length.out=100)),
           y3 = expit(1 + seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 2.5) +
  geom_line(aes(y = y2), lwd=2.5, lty = 2, color="blue") +
  geom_line(aes(y = y3), lwd=2.5, lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = -2, y = 0.4, label="1", size=8) +
  annotate("text", x = 0, y = 0.4, label="-1", size=9) +
  annotate("text", x = 3.5, y = 0.4, label="-3", size=8) +
  theme(text = element_text(size = 30))
```
]

.pull-right[
$p_i = \dfrac{\exp\{-1 +  \beta_1 \ X_i \}}{1 + \exp\{-1 +  \beta_1 \ X_i \}} \hspace{0.5cm}$ for $\beta_1 = 0.5, 1, 2$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + 0.5*seq(-5, 6, length.out=100)),
           y2 = expit(-1 + seq(-5, 6, length.out=100)),
           y3 = expit(-1 + 2*seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 2.5) +
  geom_line(aes(y = y2), lwd=2.5, lty = 2, color="blue") +
  geom_line(aes(y = y3), lwd=2.5, lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = 3, y = 0.75, label="0.5", size=8) +
  annotate("text", x = 1.5, y = 0.75, label="1", size=9) +
  annotate("text", x = 0.5, y = 0.75, label="2", size=8) +
  theme(text = element_text(size = 30))
```
]
]

---

### Interpretation

.large[
$$Y_i \sim Bernoulli(p_i)$$

$$\log \left(\dfrac{\widehat{p}_i}{1 - \widehat{p}_i}\right) = 1.737 - 0.361 \ WBC_i$$

.question[
Work in groups of 2-3 for 5 minutes on the following questions:

* Are patients with a higher WBC more or less likely to have dengue?
* What is the change in *log odds* associated with a unit increase in WBC?
* What is the change in *odds* asociated with a unit increase in WBC?
]
]

---

### Recap: ways of fitting a *linear* regression model

.large[
$$Y_i = \beta_0 + \beta_1 X_{i, 1} + \beta_2 X_{i, 2} + \cdots + \beta_k X_{i, k} + \varepsilon_i \hspace{1.5cm} \varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$$
]

.large[
.question[
Suppose we observe data $(X_1, Y_1), (X_2, Y_2), ..., (X_n, Y_n)$, where $X_i = (1, X_{i,1}, ..., X_{i,k})^T$.

How do we fit this linear regression model? That is, how do we estimate $$\beta = \begin{bmatrix} \beta_0 \\ \beta_1 \\ \vdots \\ \beta_k \end{bmatrix}$$
]
]

---

### Summary: three ways of fitting linear regression models

.large[
* Minimize SSE, via derivatives of $\sum \limits_{i=1}^n (Y_i - \beta_0 - \beta_1 X_{i,1} - \cdots - \beta_k X_{i,k})^2$
* Minimize $||Y - \widehat{Y}||$ (equivalent to minimizing SSE)
* Maximize likelihood (for *normal* data, equivalent to minimizing SSE)
]

.large[
.question[
Which of these three methods, if any, is appropriate for fitting a logistic regression model? Do any changes need to be made for the logistic regression setting?
]
]