---
title: "Lecture 10: Inference with logistic regression models"
format: 
  revealjs:
    theme: theme.scss
editor: source
editor_options: 
  chunk_output_type: console
---

## Recall: the Titanic data

```{r, include=F}
titanic <- read.csv("https://sta711-s24.github.io/homework/Titanic.csv")
```

Data on 891 passengers on the *Titanic*. Variables include:

* `Survived`
* `Pclass`
* `Sex`
* `Age`

## Logistic regression model

$$Survived_i \sim Bernoulli(p_i)$$

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Class2_i + \beta_3 Class3_i + \beta_4 Male_i + \beta_5 Age_i$$

## Fitting the model in R

```{r}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)
summary(m1)$coefficients
```

Suppose I want to know whether there is a relation between age and the probability of survival. What hypotheses would I test?

## Wald tests for single coefficients

```{r}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)
summary(m1)$coefficients
```

## Another question

```{r}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)
summary(m1)$coefficients
```

Suppose I want to know whether there is a relation between *passenger class* and the probability of survival. What hypotheses would I test?

## Recall: nested tests for linear regression

## Logistic regression model performance

```
Coefficients:
                    Estimate Std. Error z value Pr(>|z|)    
(Intercept)         3.777013   0.401123   9.416  < 2e-16 ***
as.factor(Pclass)2 -1.309799   0.278066  -4.710 2.47e-06 ***
as.factor(Pclass)3 -2.580625   0.281442  -9.169  < 2e-16 ***
Sexmale            -2.522781   0.207391 -12.164  < 2e-16 ***
Age                -0.036985   0.007656  -4.831 1.36e-06 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 964.52  on 713  degrees of freedom
Residual deviance: 647.28  on 709  degrees of freedom
  (177 observations deleted due to missingness)
AIC: 657.28

Number of Fisher Scoring iterations: 5
```

## Nested logistic regression models

```{r, echo=T}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)

m1$deviance
```

```{r, echo=T}
m2 <- glm(Survived ~ Sex + Age, 
          family = binomial, data = titanic)
m2$deviance
```

## Preview: likelihood ratio test

## Preview: likelihood ratio test

```{r, echo=T}
m1 <- glm(Survived ~ as.factor(Pclass) + Sex + Age, 
          family = binomial, data = titanic)


m2 <- glm(Survived ~ Sex + Age, 
          family = binomial, data = titanic)

pchisq(m2$deviance - m1$deviance, df=2, lower.tail=F)
```


