---
title: "Lecture 18: Wald Tests"
format: 
  revealjs:
    theme: theme.scss
editor: source
editor_options: 
  chunk_output_type: console
---


## Hypothesis test for a population proportion

Let $Y_1, Y_2,...\overset{iid}{\sim} Bernoulli(p)$. We want to test

$H_0: p = p_0 \hspace{1cm} H_A: p \neq p_0$

## Wald test for one parameter

## Testing multiple parameters

Logistic regression model for the dengue data:

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$$

Researchers want to know if there is any relationship between white blood cell count or platelet count, and the probability a patient has dengue. What hypotheses should they test?

## Testing multiple parameters

```{r include=F}
library(knitr)


dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

```{r}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)$coefficients
```

Can the researchers test their hypotheses using this output?

## Wald tests for multiple parameters


## Class activity

[https://sta711-s24.github.io/class_activities/ca_lecture_18.html](https://sta711-s24.github.io/class_activities/ca_lecture_18.html)


