---
title: "Class activity"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Convergence of random variables

In this class activity, we will use simulations to explore convergence of random variables; in particular, convergence in probability and convergence in distribution.

**Convergence in probability:** Let $X_1, X_2, ...$ be a sequence of random variables. We say that $X_n \overset{p}{\to} X$ if for all $\varepsilon > 0$,

$$\lim \limits_{n \to \infty} P(|X_n - X| \geq \varepsilon) = 0$$

**Convergence in distribution:** Let $X_1, X_2, ...$ be a sequence of random variables. We say that $X_n \overset{d}{\to} X$ if 

$$\lim \limits_{n \to \infty} F_{X_n}(t) = F_{X}(t)$$

for all points $t$ where $F_X$ is continuous; here $F_{X_n}$ is the cdf of $X_n$, and $F_X$ the cdf of $X$.

## The Weak Law of Large Numbers (WLLN)

There are multiple versions of the WLLN. Here is the version we stated in class:

**WLLN:** Let $X_1, X_2,...$ be iid with $\mu = \mathbb{E}[X_i]$ and $Var(X_i) < \infty$. Let $\bar{X}_n = \frac{1}{n} \sum \limits_{i=1}^n X_i$. Then $\bar{X}_n \overset{p}{\to} \mu$.

In this activity we will simulate data to see the WLLN in action. The code below generates 1000 samples of size $n = 20$ from a $Uniform(0, 1)$ distribution, and calculates the sample mean for each sample:

```{r, eval=F}
n <- 20
nsim <- 1000
means <- rep(NA, nsim)

for(j in 1:nsim){
  x <- runif(n)
  means[j] <- mean(x)
}
```

:::{.question}
#### Question 1

Let $\varepsilon = 0.03$. What fraction of the time do the sample means fall within $\varepsilon$ of the true mean (here $\mu = 0.5$)? That is, what is the empirical estimate of $P(|\bar{X}_n - \mu| \geq 0.03)$?
:::

:::{.question}
#### Question 2

Repeat Question 1 with different values of $n$, and plot the probability as $n$ increases. Verify that the probability approaches 0.
:::


## The Central Limit Theorem (CLT)

Here is the version of the CLT that we stated in class:

**CLT:** Let $X_1, X_2,...$ be iid random variables, whose mgf exists in a neighborhood of 0. Let $\mu = \mathbb{E}[X_i]$ and $\sigma^2 = Var(X_i) < \infty$. Then 

$$\frac{\sqrt{n}(\bar{X}_n - \mu)}{\sigma} \overset{d}{\to} Z$$

where $Z \sim N(0, 1)$.

The code below simulates 1000 samples of size $n = 2$ from a $Poisson(1)$ distribution, and calculates the sample mean for each sample. It then plots the CDF of the sample means (in black), and the CDF of a standard normal distribution (in red).

```{r, eval=F}
n <- 2
nsim <- 1000
means <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rpois(n, 1)
  means[i] <- mean(x)
}

cdf_pts <- seq(-4, 4, 0.01)
plot(cdf_pts, ecdf(sqrt(n) * (means - 1) / sqrt(1))(cdf_pts), type="l", 
     xlab = "x", ylab = "CDF")
points(cdf_pts, pnorm(cdf_pts, mean=0, sd=1), type="l", col="red")
```

:::{.question}
#### Question 3

Run the code above; is the $N(0, 1)$ distribution close to the distribution of $\bar{X}$?
:::

:::{.question}
#### Question 4

Repeat Question 3 with different values of $n$, and verify that the $N(0, 1)$ distribution is a better approximation as $n$ increases.
:::

:::{.question}
#### Question 5

Repeat Question 4 with different distributions for $X_i$, and verify that the distribution of $\bar{X}$ is still approximately normal.
:::

## Practice question

Suppose that $X_1, X_2,... \overset{iid}{\sim} Uniform(0, 1)$. Let $X_{(n)} = \max\{X_1,...,X_n\}$. Then $X_{(n)} \overset{p}{\to} 1$. 


:::{.question}
#### Question 6

Conduct a simulation to support this claim.
:::

:::{.question}
#### Question 7

Now prove the result mathematically.
:::


## Practice question

Suppose that $X_1, X_2,... \overset{iid}{\sim} Uniform(0, 1)$. Let $X_{(n)} = \max\{X_1,...,X_n\}$. Then $n(1 - X_{(n)}) \overset{d}{\to} Y$, where $Y \sim Exp(1)$. 


:::{.question}
#### Question 8

Conduct a simulation to support this claim.
:::

:::{.question}
#### Question 9

Now prove the result mathematically.
:::
