---
title: "Class Activity"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Hypothesis testing with the Titanic data

In the first part of this class activity, we will test another hypothesis about the titanic data. Let $Y_i$ denote whether passenger $i$ survived. We will use the model

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 Age_i + \beta_3 SecondClass_i + \beta_4 ThirdClass_i$$

The Titanic data can be imported into R with

```{r}
titanic <- read.csv("https://sta711-s24.github.io/homework/Titanic.csv")
```


## Contrasts

On the previous class activity, you tested the null hypothesis $H_0: \beta_3 = \beta_4 = 0$. This tests whether there is any relationship between passenger class and survival. But now suppose you want to ask a different question: is there a difference in survival between second and third class passengers? 

Neither $\beta_3$ nor $\beta_4$ directly addresses this question. Instead, we want to compare $\beta_3$ and $\beta_4$ to *each other*; our hypotheses are

$$H_0: \beta_3 - \beta_4 = 0 \hspace{1cm} H_A: \beta_3 - \beta_4 \neq 0$$

How should we test these hypotheses? Once again, we can use a Wald test!

:::{.question}
#### Question 1

Find a matrix $C$ such that $C\beta = \beta_3 - \beta_4$.
:::

The difference $\beta_3 - \beta_4$ is called a *contrast*, and is another example of a linear transformation of $\beta$. Remember that we can test hypotheses about linear transformations of $\beta$ using the Wald test. To test $H_0: C\beta = \gamma_0$, we calculate the test statistic

$$W = (C \widehat{\beta} - \gamma_0)^T(C I^{-1}(\beta) C^T)^{-1} (C \widehat{\beta} - \gamma_0)$$

Under $H_0$, $W \approx \chi^2_q$, where $q$ is the length of $C \beta$.

:::{.question}
#### Question 2

Calculate the Wald test statistic for $H_0: \beta_3 - \beta_4 = 0$. Recall that the variance-covariance matrix can be extracted with the `vcov(...)` function.
:::

:::{.question}
#### Question 3

Using your test statistic from question 2, calculate a p-value using the appropriate $\chi^2$ distribution.
:::

# Wald tests: power

In the second part of this class activity, we will use simulations to investigate the properties of a Wald test. In particular, we are interested in the probability that our test rejects the null hypothesis $H_0$.

For concreteness, consider the Wald test for a population mean which we discussed in class. Suppose that $X_1,...,X_n$ are a sample from some population with mean $\mu$, and we want to test

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu \neq \mu_0$$
Suppose we know the population variance $\sigma^2$. Our test statistic is then

$$Z_n = \frac{\sqrt{n}(\overline{X}_n - \mu_0)}{\sigma}$$

and we reject $H_0$ when $|Z_n| > z_{\alpha/2}$, where $z_{\alpha/2}$ is the upper $\alpha/2$ quantile of a $N(0, 1)$ distribution, and $\alpha$ is our desired Type I error rate. E.g., if $\alpha = 0.05$, then $z_{\alpha/2} \approx 1.96$.

## Recall: power and type I error

*Power* and *type I error* tell us about the *probability* that we reject $H_0$, under different circumstances. Formally, for the Wald test described above,

$$Power(\mu^*) = P(\text{reject } H_0 | \mu = \mu^*)$$
Ideally, we want a *high* probability of rejecting $H_0$, if $H_A$ is true. So, we want $Power(\mu^*)$ to be high for values of $\mu^* \neq \mu_0$. But if $H_0$ is true, we want to control the probability of rejecting (because a rejection would be a false positive!). We call $Power(\mu_0)$ the *type I error rate*. Typically, we want $Power(\mu_0) = \alpha$ (where we get to choose $\alpha$, and common values include $\alpha = 0.01$ or $\alpha = 0.05$).

## Simulations

For the purpose of this activity, suppose that $X_1,...,X_n \overset{iid}{\sim} N(\mu, 1)$. We wish to test

$$H_0: \mu = 0 \hspace{1cm} H_A: \mu \neq 0$$

The following code simulates data with $n = 20$ and $\mu = 0$, calculates the test statistic, and checks whether we reject:


```{r, eval=F}
alpha <- 0.05
n <- 20
mu <- 0
x <- rnorm(n, mu)
z <- sqrt(n)*(mean(x) - 0)/1

# check if reject
abs(z) > qnorm(alpha/2, lower.tail=F)
```


:::{.question}
#### Question 4

Repeat the code above many times, and verify that the fraction of times you reject $H_0$ is approximately 0.05. (This is the *type I error*, i.e. $Power(\mu_0)$)
:::

:::{.question}
#### Question 5

For fixed sample size $n = 20$, estimate $Power(\mu^*)$ for different values of $\mu^*$, and plot $Power(\mu^*)$ against $\mu^*$. How does power (our probability of rejecting) change as we change $\mu^*$?
:::

:::{.question}
#### Question 6

For fixed $\mu^* = 0.2$, estimate $Power(0.2)$ for different values of $n$, and plot $Power(0.2)$ against $n$. How does power (our probability of rejecting) change as we change $n$?
:::

For the Wald test described here, we can actually write down an approximation to the power:

$$Power(\mu^*) \approx \Phi \left( -z_{\alpha/2} - \frac{\sqrt{n}(\mu^* - \mu_0)}{\sigma} \right) + 1 - \Phi \left( z_{\alpha/2} - \frac{\sqrt{n}(\mu^* - \mu_0)}{\sigma} \right)$$

:::{.question}
#### Question 7

Using the asymptotic normality of $\overline{X}_n$, derive this result.
:::

:::{.question}
#### Question 8

Using this approximation to $Power(\mu^*)$, show that $Power(\mu^*) \to 1$ as $n \to \infty$ for any fixed $\mu^* \neq \mu_0$.
:::

:::{.question}
#### Question 9

Using this approximation to $Power(\mu^*)$, show that $Power(\mu^*) \to 1$ as $|\mu^* - \mu_0| \to \infty$ for any fixed $n$.
:::


